Data Processing: Disaster Declaration Summaries

### 2021-11-12 code/analysis_code/analysis-eda.Rmd
Finished linear models, started ML models

* There are 445 distinct disaster numbers across 2012 - 2021. 
* `OblAmt` ranges from -\$855,200.3 to \$4,408,424,297.
* `ReqAmt` ranges from -\$855,200.3 to \$4,388,540,486.
* `ResponseDays` are missing 72% of the time, which suggests it should not be included in the analysis.
* There are 16 possible incident types
* There is a clear effect of both incident month and incident year.

@@

@@

* state
* IncidentYear
* IncidentMonth
* declarationType
* incidentType
* IncidentDuration
* IH
* PA
* HM
* Population
* Counties
* TotalAgencies
* FEMACSAvg

@@

#specify table labels
table1::label(EDAdata$state) <- "State"
table1::label(EDAdata$IncidentYear) <- "Incident Year"
table1::label(EDAdata$IncidentMonth) <- "Declaration Month"
table1::label(EDAdata$declarationType) <- "Declaration Type"
table1::label(EDAdata$incidentType) <- "Incident Type"
table1::label(EDAdata$IncidentDuration) <- "Incident Duration"
table1::label(EDAdata$Population) <- "State Population"
table1::label(EDAdata$Counties) <- "Counties per State"
table1::label(EDAdata$TotalAgencies) <- "Federal Agencies Involved"
table1::label(EDAdata$FEMACSAvg) <- "Average FEMA Cost Share"
table1::label(EDAdata$IH) <- "Individuals/Households Program"
table1::label(EDAdata$PA) <- "Public Assistance Program"
table1::label(EDAdata$HM) <- "Hazard Mitigation Program"
table1::label(EDAdata$ReqAmt) <- "Requested FEMA Funds"
table1::label(EDAdata$OblAmt) <- "Obligated FEMA Funds"

#specify units
table1::units(EDAdata$IncidentDuration) <- "days"

#create table
table1 <- table1::table1(~ ReqAmt + OblAmt + state + IncidentYear + IncidentMonth + declarationType + incidentType +
                           IncidentDuration + IH + PA + HM + Population + Counties + TotalAgencies + FEMACSAvg,
                          data = EDAdata)

>>>>

* There are 445 distinct disaster numbers across 2012 - 2021
* `OblAmt` ranges from -\$855,200.3 to \$4,408,424,297
* `ReqAmt` ranges from -\$855,200.3 to \$4,388,540,486
* There are 16 possible incident types
* There is a clear effect of both incident month and incident year

@@

## Predictor: Response Duration
```{r}
#summary statistics since it's continuous
base::summary(EDAdata$ResponseDays)
#NAs are ongoing events

#density plot
ggplot2::ggplot(data = EDAdata, aes(x = ResponseDays)) +
  geom_density()

#normal boxplot
ggplot2::ggplot(data = EDAdata, aes(y = ResponseDays)) +
  geom_boxplot()

##QQ plot
car::qqPlot(EDAdata$ResponseDays)

#examine graphical relationship with outcomes of interest
req_fig_resp <- EDAdata %>%
                    ggplot2::ggplot(aes(x = ResponseDays)) +
                      geom_point(aes(y = ReqAmt),
                                 color = "tomato",
                                 show.legend = FALSE) +
                      scale_y_continuous(expand = c(0, 0),
                                         limits = c(0, 1000000000),
                                         labels = scales::dollar_format()) +
                      labs(x = "Response Duration (Days)",
                           y = "Funds ($)",
                           subtitle = "Requested")
req_fig_resp

#repeat for obligated funds
obl_fig_resp <- EDAdata %>%
                    ggplot2::ggplot(aes(x = ResponseDays)) +
                      geom_point(aes(y = OblAmt),
                                 color = "turquoise4",
                                 show.legend = FALSE) +
                      scale_y_continuous(expand = c(0, 0),
                                         limits = c(0, 1000000000),
                                         labels = scales::dollar_format()) +
                      labs(x = "Response Duration (Days)",
                           y = "Funds ($)",
                           subtitle = "Obligated")
obl_fig_resp

#combine into one plot
plot_col7 <- cowplot::plot_grid(req_fig_resp, 
                                     obl_fig_resp,
                                     ncol = 1,
                                     align = "v",
                                     rel_heights = c(0.5,0.5))
plot_col7

title8 <- cowplot::ggdraw() + 
                    draw_label(
                      "Requested vs obligated FEMA funds by response duration (Under $1B)",
                      fontface = 'bold',
                      x = 0,
                      hjust = 0) +
                    theme( plot.margin = margin(0, 0, 0, 7))

funds_fig_resp <- cowplot::plot_grid(
                              title8, plot_col7,
                              ncol = 1,
                              # rel_heights values control vertical title margins
                              rel_heights = c(0.1, 1))
funds_fig_resp

#save file
funds_resp_fig_file = here("results","funds-resp.png")
ggsave(filename = funds_resp_fig_file, plot = funds_fig_resp)
```

@@

* state
* IncidentYear
* IncidentMonth
* declarationType
* incidentType
* IncidentDuration
* ResponseDays
* IH
* PA
* HM
* Population
* Counties
* TotalAgencies
* FEMACSAvg

@@

#specify table labels
table1::label(EDAdata$state) <- "State"
table1::label(EDAdata$IncidentYear) <- "Incident Year"
table1::label(EDAdata$IncidentMonth) <- "Declaration Month"
table1::label(EDAdata$declarationType) <- "Declaration Type"
table1::label(EDAdata$incidentType) <- "Incident Type"
table1::label(EDAdata$IncidentDuration) <- "Incident Duration"
table1::label(EDAdata$ResponseDays) <- "Response Duration"
table1::label(EDAdata$Population) <- "State Population"
table1::label(EDAdata$Counties) <- "Counties per State"
table1::label(EDAdata$TotalAgencies) <- "Federal Agencies Involved"
table1::label(EDAdata$FEMACSAvg) <- "Average FEMA Cost Share"
table1::label(EDAdata$IH) <- "Individuals/Households Program"
table1::label(EDAdata$PA) <- "Public Assistance Program"
table1::label(EDAdata$HM) <- "Hazard Mitigation Program"
table1::label(EDAdata$ReqAmt) <- "Requested FEMA Funds"
table1::label(EDAdata$OblAmt) <- "Obligated FEMA Funds"

#specify units
table1::units(EDAdata$IncidentDuration) <- "days"
table1::units(EDAdata$ResponseDays) <- "days"

#create table
table1 <- table1::table1(~ ReqAmt + OblAmt + state + IncidentYear + IncidentMonth + declarationType + incidentType +
                           IncidentDuration + ResponseDays + IH + PA + HM + Population + Counties + TotalAgencies + FEMACSAvg,
                          data = EDAdata)

### 2021-11-12 code/analysis_code/analysis-linearmodels.Rmd
Finished linear models, started ML models

---
title: "Analysis: Linear Models"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---
<br>

@@

1. Fit a simple linear regression with the primary predictor of interest
2. Fit a multivariable linear regression
3. Fit a LASSO linear regression

@@

## Data Filtering
We will make two datasets with the only the relevant variables for each:

* state
* IncidentYear
* IncidentMonth
* declarationType
* incidentType
* IncidentDuration
* IH
* PA
* HM
* Population
* Counties
* TotalAgencies
* FEMACSAvg

```{r data subsets}
#obligated amount
obl_data <- analysis_data %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)

#requested amount
req_data <- analysis_data %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
```

@@

Create box and whisker plot for the fit output.
```{r}
req_SLR_bp <- broom.mixed::tidy(req_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_SLR_bp
```

@@

### Multivariable Linear Regression
Create the MVR using the `tidymodels` setup.
```{r req SLR}
#define linear mod
lr_mod <- linear_reg() %>%
              set_engine("lm")

@@

Create box and whisker plot for the fit output.
```{r}
#basic box and whisker plot
req_MVR_bp <- broom.mixed::tidy(req_MVR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_bp

@@

#box and whisker plot for lm_fit2 significant predictors
req_MVR_fit_bp2 <- req_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_fit_bp2
```

@@

Predict requested funds from FEMA  based on the model above for the training data set.
```{r req SLR predict}
#use predict to predict funds
stats::predict(req_MVR_fit, req_train_data)

@@

>>>>

---
title: "Analysis: Linear Models"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---

@@

1. Fit a simple linear regression with the primary predictor of interest
2. Fit a multivariable linear regression
3. Comparison of the two models

@@

## Data Filtering
We first need to address the NA values, and then we will make two datasets with the only the relevant variables for each:

* state
* IncidentYear
* IncidentMonth
* declarationType
* incidentType
* IncidentDuration
* ResponseDays
* IH
* PA
* HM
* Population
* Counties
* TotalAgencies
* FEMACSAvg

```{r data subsets}
#obligated amount
obl_data <- analysis_data %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)

#requested amount
req_data <- analysis_data %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)
```

@@

Create box and whisker plot for the fit output.
```{r req SLR bp}
req_SLR_bp <- broom.mixed::tidy(req_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_SLR_bp
```

@@

### Multivariable Linear Regression
Create the MVR using the `tidymodels` setup.
```{r req MVR}
#define linear mod
lr_mod <- linear_reg() %>%
              set_engine("lm")

@@

Create box and whisker plot for the fit output.
```{r req MVR bp}
#basic box and whisker plot
req_MVR_bp <- broom.mixed::tidy(req_MVR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_bp

@@

#box and whisker plot for lm_fit2 significant predictors
req_MVR_fit_bp2 <- req_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
req_MVR_fit_bp2
```
This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of requested FEMA funding. It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.

@@

Predict requested funds from FEMA  based on the model above for the training data set.
```{r req MVR predict}
#use predict to predict funds
stats::predict(req_MVR_fit, req_train_data)

@@

Assess goodness of fit measures.
```{r req MVR eval}
#using glace
modelsummary::glance(req_MVR_fit)

#using the prediction capacity
req_MVR_train %>% yardstick::metrics(truth = ReqAmt, estimate = .pred)
```
While this is an improvement, it's still not a great model. Let's compare the two side-by-side using RMSE.

### Requested Funds: Model Evaluation
Formally compare the two linear models
```{r req lm eval}
#First create comprehensive df with all of the predictions
req_lm_preds <- bind_rows(req_MVR_train,
                          req_SLR_train)

#Use the yardstick package to calculate the metrics on resamples
req_lm_metrics <- req_lm_preds %>%
                      dplyr::group_by(data, model) %>%
                      yardstick::metrics(truth = ReqAmt, estimate = .pred)

#for interpretation: rmse = Root Mean Squared Error (RMSE), rsq = R^2, mae = Mean Absolute Error (MAE)
#make a df that displays RMSE and R^2 for each model and data (all dplyr functions)
req_lm_RMSE <- req_lm_metrics %>%
                    filter(.metric == "rmse" | .metric == "rsq") %>%
                    select(-c(.estimator, data))

#save comparison table
req_lm_eval_file = here("results", "lm-req-lm-eval.Rds")
saveRDS(req_lm_RMSE, file = req_lm_eval_file)
```
In comparing the two models, the full model certainly improves the fit, but neither is ideal. Alternative models should be considered.

## Outcome of Interest: Obligated FEMA Funds
We will now examine obligated FEMA funds as the outcome of interest.

### Data Set Up
Split the data randomly into train and test datasets, and define cross validation
```{r obl data splitting}
#fix random numbers by setting the seed (helps reproducibility)
set.seed(123)

#put 3/4 of data into the training set
obl_data_split <- rsample::initial_split(obl_data, prop = 3/4)

#create dataframes for the two sets
obl_train_data <- rsample::training(obl_data_split)
obl_test_data <- rsample::testing(obl_data_split)
```

### Simple Linear Regression
Create the SLR using the `tidymodels` setup.
```{r obl SLR}
#define lr mod
lr_mod <- parsnip::linear_reg() %>%
          parsnip::set_engine("lm")

#define recipe
obl_SLR_rec <- recipes::recipe(OblAmt ~ IncidentDuration, data = obl_train_data)

#define SLR workflow
obl_SLR_wflow <- workflows::workflow() %>%
                   workflows::add_model(lr_mod) %>%
                   workflows::add_recipe(obl_SLR_rec)

#fit the training data to the workflow
obl_SLR_fit <- obl_SLR_wflow %>%
                  parsnip::fit(data = obl_train_data)

#create tibble for model fit using broom and extract
obl_SLR_fit %>%
  workflowsets::extract_fit_parsnip() %>%
  broom::tidy()
```

Create box and whisker plot for the fit output.
```{r obl SLR bp}
obl_SLR_bp <- broom.mixed::tidy(obl_SLR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
obl_SLR_bp
```

Predict requested funds from FEMA  based on the model above for the training data set.
```{r obl SLR predict}
#use predict to predict funds
stats::predict(obl_SLR_fit, obl_train_data)

#create df with model predictions and actual measures
obl_SLR_aug <- augment(obl_SLR_fit, obl_train_data)

#add labeling variables for later summarization
obl_SLR_train <- obl_SLR_aug %>%
                    dplyr::mutate(data = "train",
                                  model = "SLR")
```

Assess goodness of fit measures.
```{r obl SLR eval}
#using glace
modelsummary::glance(obl_SLR_fit)

#using the prediction capacity
obl_SLR_train %>% yardstick::metrics(truth = OblAmt, estimate = .pred)
```
This is a terrible model. Time to try adding more predictors.

### Multivariable Linear Regression
Create the MVR using the `tidymodels` setup.
```{r obl MVR}
#define linear mod
lr_mod <- linear_reg() %>%
              set_engine("lm")

#define recipe
obl_MVR_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data)

#define SLR workflow
obl_MVR_wflow <- workflows::workflow() %>%
                   workflows::add_model(lr_mod) %>%
                   workflows::add_recipe(obl_MVR_rec)

#fit the training data to the workflow
obl_MVR_fit <- obl_MVR_wflow %>%
                  parsnip::fit(data = obl_train_data)

#create tibble for model fit using broom and extract
obl_MVR_fit %>%
  workflowsets::extract_fit_parsnip() %>%
  broom::tidy()
```

Create box and whisker plot for the fit output.
```{r obl MVR bp}
#basic box and whisker plot
obl_MVR_bp <- broom.mixed::tidy(obl_MVR_fit) %>%
                dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                   whisker_args = list(color = "blue"),
                                   vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
obl_MVR_bp

#box and whisker plot for significant predictors
#first filter significant results
obl_MVR_fit_sig <- broom.mixed::tidy(obl_MVR_fit) %>%
                      dplyr::filter(p.value < 0.05)

#box and whisker plot for lm_fit2 significant predictors
obl_MVR_fit_bp2 <- obl_MVR_fit_sig %>%
                        dotwhisker::dwplot(dot_args = list(size = 2, color = "blue"),
                                           whisker_args = list(color = "blue"),
                                           vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
obl_MVR_fit_bp2
```
This model identifies total number of federal agencies, the state/territory being Puerto Rico, and emergency declarations as significant predictors of obligated FEMA funding. It makes sense that an emergency declaration is a protective factor, as it is a smaller event than a major disaster declaration.

Predict obligated funds from FEMA  based on the model above for the training data set.
```{r obl MVR predict}
#use predict to predict funds
stats::predict(obl_MVR_fit, obl_train_data)

#create df with model predictions and actual measures
obl_MVR_aug <- augment(obl_MVR_fit, obl_train_data)

#add labeling variables for later summarization
obl_MVR_train <- obl_MVR_aug %>%
                    dplyr::mutate(data = "train",
                                  model = "MVR")
```

Assess goodness of fit measures.
```{r obl MVR eval}
#using glace
modelsummary::glance(obl_MVR_fit)

#using the prediction capacity
obl_MVR_train %>% yardstick::metrics(truth = OblAmt, estimate = .pred)
```
While this is an improvement, it's still not a great model. Let's compare the two side-by-side using RMSE.

### Obligated Funds: Model Evaluation
Formally compare the two linear models
```{r obl lm eval}
#First create comprehensive df with all of the predictions
obl_lm_preds <- bind_rows(obl_MVR_train,
                          obl_SLR_train)

#Use the yardstick package to calculate the metrics on resamples
obl_lm_metrics <- obl_lm_preds %>%
                      dplyr::group_by(data, model) %>%
                      yardstick::metrics(truth = OblAmt, estimate = .pred)

#for interpretation: rmse = Root Mean Squared Error (RMSE), rsq = R^2, mae = Mean Absolute Error (MAE)
#make a df that displays RMSE and R^2 for each model and data (all dplyr functions)
obl_lm_RMSE <- obl_lm_metrics %>%
                    filter(.metric == "rmse" | .metric == "rsq") %>%
                    select(-c(.estimator, data))

#save comparison table
obl_lm_eval_file = here("results", "lm-obl-lm-eval.Rds")
saveRDS(obl_lm_RMSE, file = obl_lm_eval_file)
```
In comparing the two models, the full model certainly improves the fit, but neither is ideal. Alternative models should be considered.

## Compare Requested and Obligated Models
As one last step for the linear models, let's compare the metrics from the obligated and requested funds models.
```{r lm comp}
#add outcome labels
req_lm_RMSE$outcome <- "Requested"
obl_lm_RMSE$outcome <- "Obligated"

#combine into one table
lm_comp <- dplyr::bind_rows(req_lm_RMSE,
                            obl_lm_RMSE)

#clean up the table
lm_comp$label <- paste(lm_comp$model, lm_comp$.metric, sep=": ")

lm_comp_tbl <- lm_comp %>%
                  subset(select = -c(1:2))

lm_comp_table <- xtabs(.estimate ~ label + outcome, lm_comp_tbl)
lm_comp_table

#save file
lm_eval_file = here("results", "lm-eval.Rds")
saveRDS(lm_comp_table, file = lm_eval_file)
```
The two outcomes yield virtually identical models.

### 2021-11-12 code/processing_code/processing-declarations.Rmd
Finished linear models, started ML models

The next step is to conduct some feature engineering to create variables that represent the time elapsed between `incidentBeginDate` and `incidentEndDate` as well as `incidentBeginDate` and `disasterCloseoutDate`. These correspond to incident duration and response duration, respectively. 
```{r incident and response days}
#use `difftime` function to find the number of days 
#response days will be absurdly long, but keeping the same to maintain units
declarations_3 <- declarations_2 %>%
                    dplyr::mutate(IncidentDuration = difftime(incidentEndDate,
                                                              incidentBeginDate,
                                                              units = c("days")),
                                  ResponseDays = difftime(disasterCloseoutDate,
                                                          incidentBeginDate,
                                                          units = c("days")))
```

>>>>

The next step is to conduct some feature engineering to create variables that represent the time elapsed between `incidentBeginDate` and `incidentEndDate` as well as `incidentBeginDate` and `disasterCloseoutDate`. These correspond to incident duration and response duration, respectively. 
```{r incident and response days}
#use `difftime` function to find the number of days 
#response days will be absurdly long, but keeping the same to maintain units
declarations_3 <- declarations_2 %>%
                    dplyr::mutate(IncidentDuration = ifelse(is.na(incidentEndDate),
                                                            difftime("2021-11-10T00:00:00.000Z",
                                                                     incidentBeginDate,
                                                                     units = c("days")),
                                                            difftime(incidentEndDate,
                                                              incidentBeginDate,
                                                              units = c("days"))),
                                  ResponseDays = ifelse(is.na(disasterCloseoutDate),
                                                        difftime("2021-11-10T00:00:00.000Z",
                                                                 incidentBeginDate,
                                                                 units = c("days")),
                                                        difftime(disasterCloseoutDate,
                                                                incidentBeginDate,
                                                                units = c("days"))))
```
